Novel genetic control of migratory diapause in Australian Monarch Butterflies

### 2023-04-12 paper/paper_draft.Rmd
added a background section to the methods

To better understand how migration re-emerged in Australian monarchs, we studied continuously breeding monarch populations in Queensland. While these populations do not experience substantial seasonal fluctuations in resource availability, and thus do not migrate [@James2019], recent work has shown that reproductive diapause can be induced in some, *but not all* individuals from these populations if they are exposed to a reduction in photoperiod during development [@Freedman2018], a trait that these non-migratory populations do not display in the wild. In brief, reproductive diapause, where individuals delay their reproductive maturation, is a critical and tightly linked phenotype for migratory monarchs because it allows them to invest more heavily in lipid storage and increase their average longevity [@Herman2001]. Like other migratory behaviors, diapause is initiated by changes in day length, temperature, and other seasonal fluctuations [@Goehring2002;@iiamsPhotoperiodicClockRegulation2019].

@@

# Methods
## Background
* life history details
* why was migration lost on the islands
* have the map figure here
* why does diapause matter

@@

bias_p <- reshape2::dcast(bias, MotherID ~ likely_pheno)
bias_p$total <- bias_p$M + bias_p$R
bias_p$proportion <- bias_p$M/bias_p$total
# Figure2a <- ggplot(bias, aes(x = bias$MotherID, fill = bias$likely_pheno)) + geom_bar(position = 'dodge')
Figure2a <- ggplot(bias_p, aes(x = MotherID, y = proportion, size = total)) + geom_point() + theme_bw() + geom_hline(yintercept = .5) + xlab("Maternal Family") + ylab("Proportion Diapaused") + labs(size = "n")
Figure2al <- ggpubr::get_legend(Figure2a)
Figure2a <- Figure2a + theme(legend.position = "none")

@@

# plot
Figure2b <- plot_manhattan(dat, "gmmat_pval_likely_pheno", chr = "scaffold", significant = 0.00005, suggestive = 0.0005, log.p = T,
                           chr_order = c(paste0("chr", 1:30), "other"), highlight_style = "red") # highlight top snp
Figure2b <- Figure2b$plot + ylab("-log10(p-value)") + xlab("Position") + theme(strip.text = element_blank())
```
In order to determine the degree to which diapause status was biased between families, we constructed a Fisher's exact test for independence using 2000 simulations in R version 4.2.2 [@rcoreteamLanguageEnvironmentStatistical2022]. We then conducted a Genome-Wide Association Study (GWAS) in order to determine if any loci were significantly associated with reproductive status. Since our samples were composed of many groups of full and half-siblings and came from two genetically different years, we used the GMMAT R package to conduct a population and family structure-corrected GWAS [@Chen2020; @Chen2016]. We visualized the results by constructing a Manhattan plot using snpR [@hemstromSnpRUserFriendly2022]. We identified the genes underlying any candidate regions by hand using the MonarchBase genome browser online tool [@Zhan2013].

@@

Genetic diversity levels were extremely similar between our two years ($\pi =$ `r round(diversity$weighted_mean_pi[2], 4)` and `r round(diversity$weighted_mean_pi[3], 4)` and $H_{o}$ = `r round(diversity$weighted_mean_ho[2], 4)` and `r round(diversity$weighted_mean_ho[3], 4)` for 2016 and 2018, respectively). Each year had a slight heterozygote excess ($F_{IS}$ = `r round(diversity$weighted_mean_fis[2], 3)` and `r round(diversity$weighted_mean_fis[3],3)` for 2016 and 2018, respectively), although that excess was lower if all samples were considered jointly ($F_{IS}$ = `r round(diversity$weighted_mean_fis[1], 3)`). Pairwise $F_{ST}$ between the two years was small but marginally significant (`r round(fst$weighted_mean_fst[1], 3)`, $p =$ `r round(fst$weighted_mean_fst_p[1], 3)`). We found no significant genetic divergence between the two sample years on either the PCA or the NGSadmix results, with clustering instead driven by family, although we did find that three of our 2018 maternal families were substantially different from all other samples on our first two PC axes (Figure 1).

@@

Diapause status was strongly biased by maternal family (Fisher's Exact Test, $p < 0.001 $, see Figure 2a). We identified one small region of chromosome 11 (between positions ~5,589,000 to ~5,620,000) containing many loci associated with diapause status well beyond the background level of association, including three SNPs from two different RAD-tags that had a $p$-value  $\le 5 \times 10^{-5}$ (Figure 2b). This region overlaps the gene DPOGS208560 (LOC116768596), an E3 ubiquitin ligase, and was still strongly associated with diapause if only the 2018 butterflies alone were examined. We also identified one locus with a $p$-value  $\le 5 \times 10^{-5}$ on chromosome 16, but there were no other significantly associated loci nearby.

@@

We found that a region on Chromosome 11 containing the gene DPOGS208560 was strongly associated with diapause initiation in Monarch Butterflies in Queensland, Australia. This gene is an E3 ubiquitin ligase, a class of proteins broadly involved with proteosomal degradation of diverse substrates, including the circadian clock protein CRY2 [@xingSCFFBXL3UbiquitinLigase2013; @yooCompetingE3Ubiquitin2013], which is known to play a role in monarch photoperiodism including migratory responses to declining photoperiod [@iiamsPhotoperiodicClockRegulation2019]. Interestingly, a predicted homolog of this gene in silk moths has been previously found to be differentially expressed during larval molt-intermolt transitions [@Hu2016], which are generally controlled by fluctuations in JH and ecdysone [@Jindra2013]. JH is also known to be a major driver of reproductive development and diapause in monarchs [@Herman2001]. Taken together, it is not surprising that DPOGS208560 may participate in diapause control in monarchs.

@@

Since we observed that *only*  the DPOGS208560 gene was associated with diapause in Australian monarchs, it seems unlikely that a secondary re-introduction of migratory monarch butterflies from North America could explain the re-emergence of migratory behavior in Australia, since there would be no reason for other known migratory associated genes to have not also made the jump back across the pacific. Unfortunately, since the SNP loci that we located are not perfectly associated with diapause and are likely to only be in linkage with any causal variation rather than the causal variants themselves, our work cannot currently delineate between the other two options (maintenance of variance at DPOGS208560 during the pacific crossing or a new mutation). If the direct, causal loci were to be identified, it would then be possible to determine which gene versions are present in North America or across the Pacific and thus determine the origin of the DPOGS208560 diapause variant in Australia.

@@

```{r Figure_1_print,echo=FALSE,warning=FALSE,fig.cap="Figure 1: Principal Component Analysis of the monarchs from 2016 and 2018, colored by maternal family.",fig.width=8, fig.height=8, fig.align='center', fig.cap.style="Image Caption"}

@@

```{r Figure_2_print,echo=FALSE,warning=FALSE,fig.cap="Figure 2: Percentage of individuals in reproductive diapuase in each maternal family (A) and  GMMAT p-values for association between each SNP and diapause status across all individuals (B). SNPs with a p-value lower that 5e-5 are highlighted in red.",fig.width=8, fig.height=6, fig.align='center', fig.cap.style="Image Caption"}
grid.arrange(Figure2a + ggtitle("A"), Figure2b + ggtitle("B"), Figure2al, layout_matrix = matrix(c(1, 2, 3, NA), nrow = 2), widths = c(1, .1))

>>>>

To better understand how migration re-emerged in Australian monarchs, we studied continuously breeding monarch populations in Queensland. While these populations do not experience substantial seasonal fluctuations in resource availability, and thus do not migrate [@James2019], recent work has shown that reproductive diapause can be induced in some, *but not all* individuals from these populations if they are exposed to a reduction in photoperiod during development [@Freedman2018], a trait that these non-migratory populations do not display in the wild. In brief, reproductive diapause, where individuals delay their reproductive maturation, is a critical and tightly linked phenotype for migratory monarchs because it allows them to invest more heavily in lipid storage and increase their average longevity [@Herman2001]. Like other migratory behaviors, diapause is initiated by changes in day length, temperature, and other seasonal fluctuations [@Goehring2002;@iiamsPhotoperiodicClockRegulation2019], and is thought to be directly controlled by Juvenile Hormone (JH) concentration during development [@GreenII2019;@Herman2001].

@@

# Methods
## Background
Monarch butterflies are native to the Carribean islands; North, Central, and northern South America; and are an established but non-native in Europe, Africa, Australia, and on a range of Pacific Islands [@oplerFieldGuideWestern1999; @Zhan2014]. In North America, the species is well known for it's unique, multi-generational migratory life history wherein individuals dispersing northward in the spring do so over three to four generations and then return to their wintering grounds in a single step [@malcolmSpringRecolonizationEastern1993; @cockrellTimeTemperatureLatitudinal1993]. These wintering grounds are located either in central Mexico or the coast of California for those individuals which summer to the east or west of the Rocky mountains, respectively [@dingleDistributionMonarchButterfly2005;@urquhartAutumnalMigrationRoutes1978], although a handful also over winter in southern Florida and the Caribbean [@dockxDirectionalStabilizingSelection2007a; @dockxNORTHAMERICANMONARCH2004]. Despite summering and wintering in different locations, gene flow between the eastern and western monarch populations is apparently high enough to render them genetically homogeneous [@Hemstrom2022a; @tallaGenomicEvidenceGene2020;@lyonsLackGeneticDifferentiation2012].

In Southern Florida, the Caribbean, Central America, South America, and throughout most of their introduced range, however, monarchs are non-migratory [@dockxDirectionalStabilizingSelection2007a;@altizerPOPULATIONSMONARCHBUTTERFLIES2010;@northMonarchButterflyBiology2004]. Residency in these populations appears to be a derived trait which has arisen multiple times after the species' split from *Danaus erripus*, their closest extant relative [@Zhan2014]. The Pacific expansion of monarchs is relatively recent, with historical records indicating that they established first in Hawaii in the 1840s, likely as a result of individuals blown in during storm events that were able to survive on recently introduced milkweeds [@Zalucki2004; @Pierce2014]. Genetic evidence is consistent with an introduction to Hawaii during this time period [@Hemstrom2022a]. As described above, they then spread across the Pacific and reached Australia in approximately 1871 [@Zalucki2004], and were recorded migrating again in Australia sometime in the 1930s [see Figure 1; @James2019].

The North American migratory populations are the subject of wide-spread research on topics including but not limited to navigation and orientation [@Guerra2014; @Merlin1700; @Mouritsen10162], parasite ecology [@satterfieldMigratoryMonarchsThat2018], and life-history associated phenotypic variation [@Herman2001; @Zhan2014; @Zhan2011; @Dingle1977], and are differentiated from non-migratory individuals by a range of phenotypic variables such as reproductive diapause, wing shape, and other physiological factors [@altizerPOPULATIONSMONARCHBUTTERFLIES2010; @hermanSTUDIESADULTREPRODUCTIVE1981; @browerFuelingFallMigration2006]. Note, however, that the monarch migratory syndrome is not necessarily genetically linked and can become split apart by changes in selection pressures [@Tenger-Trolander2019]. The migratory populations outside of North America, however, are not yet heavily studied, and the mechanisms enabling migration in them are not yet well understood.

@@

bias_p <- reshape2::dcast(bias, MotherID ~ likely_pheno)
bias_p$total <- bias_p$M + bias_p$R
bias_p$proportion <- bias_p$M/bias_p$total
# Figure3a <- ggplot(bias, aes(x = bias$MotherID, fill = bias$likely_pheno)) + geom_bar(position = 'dodge')
Figure3a <- ggplot(bias_p, aes(x = MotherID, y = proportion, size = total)) + geom_point() + theme_bw() + geom_hline(yintercept = .5) + xlab("Maternal Family") + ylab("Proportion Diapaused") + labs(size = "n")
Figure3al <- ggpubr::get_legend(Figure3a)
Figure3a <- Figure3a + theme(legend.position = "none")

@@

# plot
Figure3b <- plot_manhattan(dat, "gmmat_pval_likely_pheno", chr = "scaffold", significant = 0.00005, suggestive = 0.0005, log.p = T,
                           chr_order = c(paste0("chr", 1:30), "other"), highlight_style = "red") # highlight top snp
Figure3b <- Figure3b$plot + ylab("-log10(p-value)") + xlab("Position") + theme(strip.text = element_blank())
```
In order to determine the degree to which diapause status was biased between families, we constructed a Fisher's exact test for independence using 2000 simulations in R version 4.2.2 [@rcoreteamLanguageEnvironmentStatistical2022]. We then conducted a Genome-Wide Association Study (GWAS) in order to determine if any loci were significantly associated with reproductive status. Since our samples were composed of many groups of full and half-siblings and came from two genetically different years, we used the GMMAT R package to conduct a population and family structure-corrected GWAS [@Chen2020; @Chen2016]. We visualized the results by constructing a Manhattan plot using snpR [@hemstromSnpRUserFriendly2022]. We identified the genes underlying any candidate regions by hand using the MonarchBase genome browser online tool [@Zhan2013].

@@

Genetic diversity levels were extremely similar between our two years ($\pi =$ `r round(diversity$weighted_mean_pi[2], 4)` and `r round(diversity$weighted_mean_pi[3], 4)` and $H_{o}$ = `r round(diversity$weighted_mean_ho[2], 4)` and `r round(diversity$weighted_mean_ho[3], 4)` for 2016 and 2018, respectively). Each year had a slight heterozygote excess ($F_{IS}$ = `r round(diversity$weighted_mean_fis[2], 3)` and `r round(diversity$weighted_mean_fis[3],3)` for 2016 and 2018, respectively), although that excess was lower if all samples were considered jointly ($F_{IS}$ = `r round(diversity$weighted_mean_fis[1], 3)`). Pairwise $F_{ST}$ between the two years was small but marginally significant (`r round(fst$weighted_mean_fst[1], 3)`, $p =$ `r round(fst$weighted_mean_fst_p[1], 3)`). We found no significant genetic divergence between the two sample years on either the PCA or the NGSadmix results, with clustering instead driven by family, although we did find that three of our 2018 maternal families were substantially different from all other samples on our first two PC axes (Figure 2).

@@

Diapause status was strongly biased by maternal family (Fisher's Exact Test, $p < 0.001 $, see Figure 3a). We identified one small region of chromosome 11 (between positions ~5,589,000 to ~5,620,000) containing many loci associated with diapause status well beyond the background level of association, including three SNPs from two different RAD-tags that had a $p$-value  $\le 5 \times 10^{-5}$ (Figure 3b). This region overlaps the gene DPOGS208560 (LOC116768596), an E3 ubiquitin ligase, and was still strongly associated with diapause if only the 2018 butterflies alone were examined. We also identified one locus with a $p$-value  $\le 5 \times 10^{-5}$ on chromosome 16, but there were no other significantly associated loci nearby.

@@

We found that a region on Chromosome 11 containing the gene DPOGS208560 was strongly associated with diapause initiation in Monarch Butterflies in Queensland, Australia. This gene is an E3 ubiquitin ligase, a class of proteins broadly involved with proteosomal degradation of diverse substrates, including the circadian clock protein CRY2 [@xingSCFFBXL3UbiquitinLigase2013; @yooCompetingE3Ubiquitin2013], which is known to play a role in monarch photoperiodism including migratory responses to declining photoperiod [@iiamsPhotoperiodicClockRegulation2019]. Interestingly, a predicted homolog of this gene in silk moths has been previously found to be differentially expressed during larval molt-intermolt transitions [@Hu2016], which are generally controlled by fluctuations in JH and ecdysone [@Jindra2013]. JH is also known to be a major driver of reproductive development and diapause in monarchs [@Herman2001;@GreenII2019]. Taken together, it is not surprising that DPOGS208560 may participate in diapause control in monarchs.

@@

Since we observed that *only*  the DPOGS208560 gene was associated with diapause in Australian monarchs, it seems unlikely that a secondary re-introduction of migratory monarch butterflies from North America could explain the re-emergence of migratory behavior in Australia, since there would be no reason for other known migratory associated genes to have not also made the jump back across the pacific. Unfortunately, since the SNP loci that we located are not perfectly associated with diapause and are likely to only be in linkage with any causal variation rather than the causal variants themselves, our work cannot currently delineate between the other two options (maintenance of variance at DPOGS208560 during the Pacific crossing or a new mutation). If the direct, causal loci were to be identified, it would then be possible to determine which gene versions are present in North America or across the Pacific and thus determine the origin of the DPOGS208560 diapause variant in Australia.

@@

```{r Figure_2_print,echo=FALSE,warning=FALSE,fig.cap="Figure 2: Principal Component Analysis of the monarchs from 2016 and 2018, colored by maternal family.",fig.width=8, fig.height=8, fig.align='center', fig.cap.style="Image Caption"}

@@

```{r Figure_3_print,echo=FALSE,warning=FALSE,fig.cap="Figure 3: Percentage of individuals in reproductive diapuase in each maternal family (A) and  GMMAT p-values for association between each SNP and diapause status across all individuals (B). SNPs with a p-value lower that 5e-5 are highlighted in red.",fig.width=8, fig.height=6, fig.align='center', fig.cap.style="Image Caption"}
grid.arrange(Figure3a + ggtitle("A"), Figure3b + ggtitle("B"), Figure3al, layout_matrix = matrix(c(1, 2, 3, NA), nrow = 2), widths = c(1, .1))
